### 2009-06-02 paper.tex
More edits

The four types of allowed transitions listed above can be formalized
as follows.
Let the total order on the nodes correspond to any preorder traversal of
the tree; thus, ``$m$ is ancestral to $n$'' is sufficient-but-not-necessary for ``$m \prec n$''.
Let ${\cal T}_m$ denote the singlet or branch transducer which emits symbols to node $m$.
Transducer ${\cal T}_m$ changes state if and only if one of the
following three mutually-exclusive conditions holds:
\begin{description}
\item{Type 1:} Transducer ${\cal T}_m$ is not in a $\Swait$ state,
  while all its successor transducers ${\cal T}_n$ are in $\Swait$
  states (where $m \prec n$).  ${\cal T}_m$ is free to make any transition.
\item{Type 2:} Transducer ${\cal T}_m$ is in a $\Swait$ state. Its parent transducer enters a $\Smatch$ or $\Sinsert$ state, emitting a symbol and forcing ${\cal T}_m$ into a $\Smatch$ state so it can absorb that symbol.
\item{Type 3:} Transducer ${\cal T}_m$ is in a $\Swait$ state. Its parent transducer enters the $\Send$ state, forcing ${\cal T}_m$ into the $\Send$ state as well.
\end{description}

@@

We implemented our model construction algorithm on the three-taxon
star phylogeny.
Given a singlet transducer modeling ancestral structures and a branch
transducer modeling structural evolution, our Perl modules generate
C++ code for the corresponding jointly-normalized three-sequence
(Triplet) SCFG.
Any model of structural evolution which can be represented as a Pair
SCFG and factored into singlet and branch transducers is permitted as input to
the packages, allowing for flexible, automated model design.

The available software is described in \supptext{3}.

@@

We illustrated our method for building models of structured sequences
using a simplified model which was introduced in previous work,
the TKF Structure Tree \cite{Holmes2004}, a simplified probabilistic
model of the evolution of RNA structure.

@@

We parametrized the singlet and branch transducers of the TKFST model
using estimates reported by a phylo-grammar for RNA secondary
structure prediction, \pfold\ \cite{KnudsenHein2003} and an
implementation of pairwise alignment for the TKF Structure Tree model,
\evoldoer\ \cite{Holmes2004}.
The equilibrium distributions of unpaired and paired nucleotides of
the singlet and branch transducers, as well as the substitution models
of unpaired and paired nucleotides of the branch transducers, were
derived from the substitution rate matrices of the \pfold\ program.
These rate matrices, which have proven useful for RNA structure
prediction \cite{KnudsenHein2003,DowellEddy2004,BradleyEtAl2008},
were derived from the Bayreuth tRNA database \cite{SprinzlEtAl1998}
and the European large subunit rRNA database \cite{DeRijkEtAl1998}.

@@

We used our model-construction algorithm to build the grammar
corresponding to the TKFST model acting on a star phylogeny with three
(extant) leaf sequences and a single (unobserved) ancestral
sequence. We chose this phylogeny for two reasons: (1) it is the
simplest extension of the well-studied, standard two-sequence (Pair
SCFG) model and (2) algorithms on a phylogeny with three leaves
should be sufficient for ergodic sampling of reconstructions on any larger phylogeny,
using e.g. a Gibbs-sampling MCMC kernel \cite{JensenHein2002}
or a progressive suboptimal-alignment sampling heuristic \cite{PatenHolmesBirney2008}.

The statistics of the TKFST model on the three-taxon phylogeny
illustrate the advantages of our procedure for model construction.
While the singlet and branch transducers are relatively simple---the
singlet transducer, shown in \tabref{tkfstsinglet}, has 7 total states
and 2 bifurcation states and the branch transducer, shown in
\tabref{tkfstbranch}, has 21 total states and 6 bifurcation
states---the ensemble model of three extant sequences is very complex.
The naive exponential upper-bound gives a maximal state space of size
$O(7 \cdot 21^{3}) \Rightarrow 6 \cdot 10^4$ states.  Using our
uninformed search algorithm, we determined that there are 287
accessible states and 686 possible transitions between these states
(compare with the $287^2 \simeq 8 \cdot 10^4$ transitions estimated
with the exponential calculation).  After performing the
transformations described in \secname{Exact elimination of null cycles in SCFGs} to
eliminate useless windback states, the ensemble model has a reduced
state space with 230 states, albeit at the cost of extra transitions,
bring the total to 1,789 transitions (here we are trading reduced
memory complexity, which is linear in the number of states, for
increased time complexity, which is linear in the number of
transitions).  Note that both before and after the reduction in
complexity, the total number of states and transitions are less than
the approximate bounds of $O(a \cdot b \cdot N) = O(7 \cdot 21 \cdot
3) \Rightarrow 441$ states and $O(a \cdot b^2 \cdot N) = O(7 \cdot
21^2 \cdot 3) \Rightarrow 9,261$ transitions suggested in \secname{The composition algorithm}.
 Nonetheless, the extreme complexity of the
ensemble model, despite the simplicity of the underlying model of RNA
structure, makes clear the necessity for automated procedures for
model construction.  Dataset S1 gives the state space of the ensemble
model constructed by the search algorithm and Dataset S2 the reduced
model after eliminating windback states; both are in Graphviz format for
visualization and show the state of the singlet transducer generating
ancestral sequence as well as the states of the branch transducers
generating observed sequences.

@@

The inference algorithms in \indiegram\ could be further constrained
to enforce, for example, a fixed multiple alignment or a consensus
structure for extant sequences.  While experimentally-determined
structures of individual RNAs are relatively rare, curated deep sequence
alignments, such as those constructed for ribosomal RNAs
\cite{Gutell93}, are frequently available for characterized RNA families.  By constraining the
inference algorithms with such sequence alignments, the memory and
time complexity of the algorithms could be dramatically reduced.  Such
constraints can be naturally expressed with {\em alignment envelopes},'
the alignment-space analogue of fold envelopes \cite{Holmes2005}.
However, in this paper we focus on model construction and inference
algorithms and postpone exploration of heuristics and constraints of
these algorithms for future work.

@@

\newpage
\section*{Discussion}

@@

An alternative MCMC scheme for sampling RNA phylogeny, structure and
alignment was developed for the SimulFold program
\cite{MeyerMiklos2007}.  SimulFold does not use a strictly normalized
probabilistic model, resulting in some oddities in the ways that
structure and indels interact (for example, it does not penalize
deletion of one half of a basepair).  Currently, it is not clear how
appropriate SimulFold would be for ancestral reconstruction, although
it has several advantages (e.g. explicit treatment of pseudoknots).
Of course, MCMC kernels are inherently adaptable to other purposes:
the MCMC moves developed for SimulFold may be useful for inference
under different models.

This paper focuses on the case where the tree topology is known, but
many of the methods which we describe below can be extended to the
more general case where none of the possible constraints (phylogeny,
structure or alignment) are final.  For example, the probabilistic
framework readily allows us to compare likelihoods of two different
phylogenetic trees by constructing a composite transducer for each
tree.  Thus, the MCMC samplers described above for alignments could,
in principal, be extended to phylogenies (albeit at a computational
cost).

>>>>

The four types of allowed transitions listed above can be formalized
as follows.
Let the total order on the nodes correspond to any preorder traversal of
the tree; thus, ``$m$ is ancestral to $n$'' is sufficient-but-not-necessary for ``$m \prec n$.''
Let ${\cal T}_m$ denote the singlet or branch transducer which emits symbols to node $m$.
Transducer ${\cal T}_m$ changes state if and only if one of the
following three mutually-exclusive conditions holds:
\begin{description}
\item{Type 1:} Transducer ${\cal T}_m$ is not in a $\Swait$ state,
  while all its successor transducers ${\cal T}_n$ are in $\Swait$
  states (where $m \prec n$).  ${\cal T}_m$ is free to make any transition.
\item{Type 2:} Transducer ${\cal T}_m$ is in a $\Swait$ state. Its parent transducer enters a $\Smatch$ or $\Sinsert$ state, emitting a symbol and forcing ${\cal T}_m$ into a $\Smatch$ state so it can absorb that symbol.
\item{Type 3:} Transducer ${\cal T}_m$ is in a $\Swait$ state. Its parent transducer enters the $\Send$ state, forcing ${\cal T}_m$ into the $\Send$ state as well.
\end{description}

@@

We implemented our model construction algorithm on the three-taxon
star phylogeny.
Given a singlet transducer modeling ancestral structures and a branch
transducer modeling structural evolution, our Perl modules generate
C++ code for the corresponding jointly-normalized three-sequence
(Triplet) SCFG.
Any model of structural evolution which can be represented as a Pair
SCFG and factored into singlet and branch transducers is permitted as input to
the packages, allowing for flexible, automated model design.
The available software is described in \supptext{3}.

@@

We illustrated our method for building models of structured sequences
using a model which was introduced in previous work,
the TKF Structure Tree \cite{Holmes2004}, a simplified probabilistic
model of the evolution of RNA structure.

@@

We parametrized the singlet and branch transducers of the TKFST model
using estimates reported by a phylo-grammar for RNA secondary
structure prediction, \pfold\ \cite{KnudsenHein2003}, and an
implementation of pairwise alignment for the TKF Structure Tree model,
\evoldoer\ \cite{Holmes2004}.
The equilibrium distributions of unpaired and paired nucleotides of
the singlet and branch transducers, as well as the substitution models
of unpaired and paired nucleotides of the branch transducers, were
derived from the substitution rate matrices of the \pfold\ program.
These rate matrices, which have proven useful for RNA structure
prediction \cite{KnudsenHein2003,DowellEddy2004,BradleyEtAl2008},
were derived from the Bayreuth tRNA database \cite{SprinzlEtAl1998}
and the European large subunit rRNA database \cite{DeRijkEtAl1998}.

@@

We used our model-construction algorithm to build the grammar
corresponding to the TKFST model acting on a star phylogeny with three
(extant) leaf sequences and a single (unobserved) ancestral
sequence. We chose this phylogeny for two reasons: (1) it is the
simplest extension of the well-studied, standard two-sequence (Pair
SCFG) model and (2) algorithms on a phylogeny with three leaves
should be sufficient for ergodic sampling of reconstructions on any larger phylogeny,
using, e.g., a Gibbs-sampling MCMC kernel \cite{JensenHein2002}
or a progressive suboptimal-alignment sampling heuristic \cite{PatenHolmesBirney2008}.

The statistics of the TKFST model on the three-taxon phylogeny
illustrate the advantages of our procedure for model construction.
While the singlet and branch transducers are relatively simple---the
singlet transducer, shown in \tabref{tkfstsinglet}, has 7 total states
and 2 bifurcation states and the branch transducer, shown in
\tabref{tkfstbranch}, has 21 total states and 6 bifurcation
states---the ensemble model of three extant sequences is very complex.
The naive exponential upper-bound gives a maximal state space of size
$O(7 \cdot 21^{3}) \Rightarrow 6 \cdot 10^4$ states.  Using our
uninformed search algorithm, we determined that there are 287
accessible states and 686 possible transitions between these states
(compare with the $287^2 \simeq 8 \cdot 10^4$ transitions estimated
with the exponential calculation).  After performing the
transformations described in \secname{Exact elimination of null cycles in SCFGs} to
eliminate useless windback states, the ensemble model has a reduced
state space with 230 states, albeit at the cost of extra transitions,
bring the total to 1,789 transitions (here we are trading reduced
memory complexity, which is linear in the number of states, for
increased time complexity, which is linear in the number of
transitions).  Note that both before and after the reduction in
complexity, the total number of states and transitions are less than
the approximate bounds of $O(a \cdot b \cdot N) = O(7 \cdot 21 \cdot
3) \Rightarrow 441$ states and $O(a \cdot b^2 \cdot N) = O(7 \cdot
21^2 \cdot 3) \Rightarrow 9,261$ transitions suggested in \secname{The composition algorithm.}
 Nonetheless, the extreme complexity of the
ensemble model, despite the simplicity of the underlying model of RNA
structure, makes clear the necessity for automated procedures for
model construction.  Dataset S1 gives the state space of the ensemble
model constructed by the search algorithm and Dataset S2 the reduced
model after eliminating windback states; both are in Graphviz format for
visualization and show the state of the singlet transducer generating
ancestral sequence as well as the states of the branch transducers
generating observed sequences.

@@

The inference algorithms in \indiegram\ could be further constrained
to enforce, for example, a fixed multiple alignment or a consensus
structure for extant sequences.  While experimentally-determined
structures of individual RNAs are relatively rare, curated deep sequence
alignments, such as those constructed for ribosomal RNAs
\cite{Gutell93}, are frequently available for characterized RNA families.  By constraining the
inference algorithms with such sequence alignments, the memory and
time complexity of the algorithms could be dramatically reduced.  Such
constraints can be naturally expressed with ``alignment envelopes,''
the alignment-space analogue of fold envelopes \cite{Holmes2005}.
However, in this paper we focus on model construction and inference
algorithms and postpone exploration of heuristics and constraints of
these algorithms for future work.

@@

\section*{Discussion}

@@

An alternative MCMC scheme for sampling RNA phylogeny, structure and
alignment was developed for the SimulFold program
\cite{MeyerMiklos2007}.  SimulFold does not use a strictly normalized
probabilistic model, resulting in some oddities in the ways that
structure and indels interact (for example, it does not penalize
deletion of one half of a basepair).  Currently, it is not clear how
appropriate SimulFold would be for ancestral reconstruction, although
it has several advantages (e.g., explicit treatment of pseudoknots).
Of course, MCMC kernels are inherently adaptable to other purposes:
the MCMC moves developed for SimulFold may be useful for inference
under different models.

This paper focuses on the case where the tree topology is known, but
many of the methods which we have described can be extended to the
more general case where none of the possible constraints (phylogeny,
structure or alignment) are final.  For example, the probabilistic
framework readily allows us to compare likelihoods of two different
phylogenetic trees by constructing a composite transducer for each
tree.  Thus, the MCMC samplers described above for alignments could,
in principal, be extended to phylogenies (albeit at a computational
cost).
